Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-12 paper-new/discussion.tex
25 pages

A line of work by New, Licata and Ahmed proves graduality by
interpreting type precision $c : A \ltdyn A'$ as an
\emph{embedding-projection pair}, that is a pure function $e : A \to
A'$ and a possibly erroring projection $p : \li A' \multimap \li A$
such that $p \circ e = \id$ and $e \circ p \ltdyn \id$
\cite{new-ahmed2018,new-licata18,new-licata-ahmed2019}. At first glance, our model
of precision looks somewhat different: it is a \emph{relation}
$\sem{c} : A \rel A'$ with a quasi-representability structure that
gives us something like $e$ and $p$ in an embedding-projection
pair. The relationship between these models is that \emph{if} the
relation were \emph{truly} representable rather than only
quasi-representable we would have that $e$ and $p$ form a \emph{Galois
connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
e$. However, we have dropped the stronger property of retraction from
our analysis in this work. With true representability, the relation
$c$ is \emph{uniquely determined} by the embedding $e$, but since we
only have quasi-representability, we need to keep the relation around
explicitly. So the extra complexity of managing explicit relations is
a cost of the intensional reasoning that guarded type theory
introduces.

@@

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We have formalized the major components of the definition of the concrete model
described in Section \ref{sec:concrete-relational-model}: predomains, error
domains, their morphisms, relations, and squares, and we used the guarded
features to define the free error domain, the lock-step error ordering, and weak
bisimilarity. We also formalized the no-go theorem from Section
\ref{sec:concrete-term-model}.

>>>>

A line of work by New, Licata and Ahmed proves graduality by
interpreting type precision $c : A \ltdyn A'$ as an
\emph{embedding-projection pair}, that is a pure function $e : A \to
A'$ and a possibly erroring projection $p : \li A' \multimap \li A$
such that $p \circ e = \id$ and $e \circ p \ltdyn \id$
\cite{new-ahmed2018,new-licata18,new-licata-ahmed2019}. At first glance, our model
of precision looks different: it is a \emph{relation}
$\sem{c} : A \rel A'$ with a quasi-representability structure that
gives us something like $e$ and $p$ in an embedding-projection
pair. The relationship between these models is that \emph{if} the
relation were \emph{truly} representable we would have that $e$ and $p$ form a \emph{Galois
connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
e$. However, we have dropped the stronger property of retraction from
our analysis in this work. With true representability, the relation
$c$ is \emph{uniquely determined} by the embedding $e$, but since we
only have quasi-representability, we need to keep the relation around
explicitly. So the extra complexity of managing explicit relations is
a cost of the intensional reasoning that guarded type theory
introduces.

@@

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We formalized the major components of the definition of the concrete model
described in Section \ref{sec:concrete-relational-model}: predomains, error
domains, morphisms, relations, squares, using the guarded
features to define the free error domain, the lock-step error ordering, and weak
bisimilarity. We also formalized the no-go theorem from Section
\ref{sec:concrete-term-model}.
